---
title: "RXLR_Rmarkdown_template"
output: html_document
date: "2023-08-30"
---


## Before you begin
### 1. Pull Github repository to local machine https://github.com/grunwaldlab/oomy_RXLRs
### 2. If working within oomy_RXLRs folder, add your FASTA and GFF files to the 'data' folder
### 3. If creating a new directory, make a 'data' subdirectory and copy your FASTA and GFF files to new directory. Second, copy the scripts folder from 'oomy_RXLRs'so you also have a 'scripts' subdirectory. Don't manipulate this folder.
### 4. If running code chunks using Rmarkdown template, copy the Rmarkdown template into your new directory as well 
### 5. Start a new R project in the same directory as your Rmarkdown template-this is important or else some scripts will not be found and cannot load. 
### 6. Install programs and dependencies described in the book chapter. If programs do not run, you may not have specified the paths to the executables, you may have to add the absolute path to the executables within the code chunks. 
### 7. Some programs allow you to specify number of cores. Please change based on your own computing capabilities. We recommend having at least 4 cores. Programs will take longer depending on genome size. 

## Setup R environment once you've installed other programs and dependencies
```{r setup, include=TRUE}
# In R environment
knitr::opts_chunk$set(echo = TRUE)
source("scripts/secretion_funcs.R")
library("tidyverse")
library("seqinr")
devtools::install_github(repo = "grunwaldlab/effectR", build_vignettes = TRUE)
library("effectR")
library("optparse")
```

## SECTION 3.2-Obtain candidate RXLRs from all ORFs
### Step 1, step 2 and first part of step 3 
### Example dataset is used (found within in 'data' folder on Github repo)
### Bash code chunk
```{bash, eval = TRUE}
# In Bash environment
#Step 1
mkdir output_data

genome="data/Paga_3770v2_chr10.fasta" #Change depending on name of input fasta assembly file

isolate=$(basename -s ".fasta" ${genome})
minsize_aa=70 # amino acids
minsize_nt=210 # minsize_aa*3 = number of nucleotides
orfs_name="${isolate}.orfs-min${minsize_aa}long.start2stop"
orfs="output_data/${orfs_name}.fasta"

# Plug parameters into EMBOSS GetORF
# Parameter find=1 sets search to between start and stop codons
getorf -sequence "$genome" -outseq "$orfs" -minsize $minsize_nt --find 1

# Step 2
# May need to specify path of tmhmm
# Had to copy signalp to /usr/bin; this may or may not be required

threads=4 #Change based on computing resources available
python2.7 scripts/rxlr_signalpeptide.py $orfs $threads \
  secretome output_data/${orfs_name}.rxlr_signalpep.out
  
# May need to specify absolute path for tmhmm
tmhmm -short $orfs > output_data/${orfs_name}.tmhmm.out 

# First part of step 3 
hmmsearch --cpu 4 --seed 123 -T 0 --tblout output_data/${orfs_name}.rxlr_WYfold_hmm.out data/WY_fold.hmm $orfs > output_data/${orfs_name}.rxlr_WYfold_hmm.log
# simple list of candidates
egrep -v '^#' output_data/${orfs_name}.rxlr_WYfold_hmm.out | awk '{print $1}' > output_data/${orfs_name}.rxlr_WYfold_hmm.list
```

### Step 3, part 2 continued
### R code chunk
```{r eval = TRUE}
# In R environment
wy_list_regex <- "*.orfs-min70long.start2stop.rxlr_WYfold_hmm.list"
orfs_wys <-  list.files(path = "output_data/",
                                pattern = wy_list_regex, full.names = TRUE) %>%
  map_dfr(read_wy_list) %>%
  unite("ID_isolate", ID, isolate, sep=":", remove = FALSE) %>%
  mutate("method" = "WY_hmm")
```

### Step 3, part 3 continued 
### R code chunk
```{r eval = TRUE}
# In R environment
# Locate files programmatically in R environment
# all ORFs:
orf_fs_regex <- ".*orfs-min70long.start2stop.fasta" 
orf_fs <- list.files(path = "output_data", 
                                pattern = orf_fs_regex, full.names = TRUE) %>% 
  tibble("orf_fs" = .) %>% 
  separate(orf_fs, into = c("isolate", NA), 
           sep = "(?=\\.orfs-min)", remove = FALSE) %>% 
  mutate(isolate = str_remove(isolate, ".*data/*")) 

# SignalP output files: 

sp3_fs_regex <- ".*orfs-min70long.start2stop.rxlr_signalpep.outsp3_tabular.tmp" 
sp3_fs <- list.files(path = "output_data", 

                                pattern = sp3_fs_regex, full.names = TRUE) %>% 
  tibble("sp3_fs" = .) %>% 
  separate(sp3_fs, into = c("isolate", NA), 
           sep = "(?=\\.orfs-min)", remove = FALSE) %>% 
  mutate(isolate = str_remove(isolate, ".*data/*")) 

orf_sp3_fs <- full_join(orf_fs, sp3_fs, by = "isolate") %>% 
  select(orf_fs, sp3_fs, isolate) 

# run motif search 
allorfs_re_effectr <- map2(orf_sp3_fs$orf_fs, orf_sp3_fs$sp3_fs, effectr_eer_sp3)

allorfs_re_effectr_tb <- tibble(bind_rows(allorfs_re_effectr)) %>%
  unite("ID_isolate", ID, isolate, sep=":", remove = FALSE)
```

### Step 4
### R code chunk
```{r eval = TRUE}
# In R environment 
# Programmatically find files in R environment 
sp3_fs_regex <- ".*orfs-min70long.start2stop.rxlr_signalpep.outsp3_tabular.tmp" 
sp3_fs <- list.files(path = "output_data", 
                                pattern = sp3_fs_regex, full.names = TRUE) 
tmm_fs_regex <- ".*orfs-min70long.start2stop.tmhmm.out" 
tmm_fs <- list.files(path= "output_data", 
                                pattern = tmm_fs_regex, full.names = TRUE) 
 
# make dataframe for easier function running 
# and validate the signalp and tmhmm runs line up 
sp3_tmm_fs <- tibble("sp3_fs" = sp3_fs, 
                     "tmm_fs " = tmm_fs) %>% 
  mutate("isolate_s" = str_extract(sp3_fs, "(?<=/).*(?=\\.orfs)")) %>% 
  mutate("isolate_t" = str_extract(tmm_fs, "(?<=/).*(?=\\.orfs)")) 
 
# parse signalP and tmhmm output files. 
# filters based on HMM_Sprob_score of signalp 
# and position of helix relative to most likely signal peptide position 

sp3_tmm_pass <- pmap(sp3_tmm_fs, ~ read_prune_tms(..1, ..2, ..3), .id = 'isolate') %>% 
  bind_rows() %>% 
  unite("ID_isolate", Protein, isolate, sep= ":", remove = FALSE) 
 
# Size of complete secretome 
sp3_tmm_pass %>% 
  filter(HMM_Sprob_score >= 0.9) %>% 
  distinct(ID_isolate, .keep_all = TRUE) %>% 
  group_by(isolate) %>% 
  summarize(n_noTM_yesSP = n()) 
```

### Step 5
### R code chunk
```{r eval = TRUE}
# In R environment 
# Start with results of RXLR-EER domain searches 

rxlr_orfs <- allorfs_re_effectr_tb %>%  
  bind_rows(orfs_wys) %>% # join to results of WY searching 
  group_by(ID_isolate, ID, isolate) %>% 
  mutate(isolate = str_remove(isolate, "/")) %>% 
  # I want the methods as a list 
  summarize(methods_list = paste(method, collapse = ",")) %>% 
  filter((grepl("Whis2007", methods_list)) | 
           (grepl("Whis_rxlr", methods_list) & (grepl("Whis_eer", methods_list) | grepl("WY_hmm", methods_list))) | 
           (grepl("Win2007", methods_list) & (grepl("Whis_eer", methods_list) | grepl("WY_hmm", methods_list)))) %>% 
  filter(ID_isolate %in% sp3_tmm_pass$ID_isolate) %>% # filter to secretome 
  group_by(isolate) %>% 
  distinct(ID_isolate) 

# Final counts 
rxlr_orfs %>%
  summarize(noTM_whis_or_whiswinrxlrEER_whiswinrxlrWY_count = n())

out_prefix <- "/orfs_cand_RXLRs_"
# Write final candidates to lists using respective isolate IDs 
allorfs_re_effectr_tb %>%
  ungroup() %>%
  distinct(ID_isolate, .keep_all = TRUE) %>%
  select(ID, isolate) %>%
  group_by(isolate) %>%
  group_walk(~ writeLines(.x$ID, paste0("output_data", out_prefix, .y$isolate, ".txt")))
```

### Step 6
### Bash code chunk
```{bash eval = TRUE}
# In Bash environment
# If AGAT was installed in a conda environment, may need to activate it
eval "$(conda shell.bash hook)"
conda activate agat

# In bash environment, change directory names if needed 
outdir="output_data" 

# This must be changed depending on location of user’s data 
assembly_dir="data"  
rxlr_prefix="orfs_cand_RXLRs_" 

# Run on all genomes being annotated 
for orf_rxlr_list in $(ls -1 $outdir/${rxlr_prefix}*.txt); do
  # Establish output file name conventions
  list_name=$(basename --suffix ".txt" $orf_rxlr_list)
  iso_name=$(echo $list_name | sed "s/$rxlr_prefix//")
  assembly=$(ls ${assembly_dir}/${iso_name}.fasta) || \
    assembly=$(ls ${assembly_dir}/other_refs/${iso_name}.fasta)

  # If needed, replace 70 with minimum protein length chosen.
  orfs="$outdir/${iso_name}.orfs-min70long.start2stop.fasta"

  # Obtain nationality. First get long ORF names from abbreviated.
  grep -w -f $orf_rxlr_list $orfs | sed 's/>//' > $outdir/${list_name}_longnames.list
  # Next convert ORF tag list to GFF
  Rscript scripts/getorf_seqnames2gff.R -i $outdir/${list_name}_longnames.list

  # Notifies about overlapping genes
  agat_sp_fix_overlaping_genes.pl -f $outdir/${list_name}_longnames.gff \
  -o $outdir/${list_name}_longnames.merge_ovlp.gff
  
  # If genes need merging, pick ORFs by hand or use output from above
  agat_sp_add_start_and_stop.pl --gff $outdir/${list_name}_longnames.gff \
  --fasta $assembly \
  --output $outdir/${list_name}_longnames.str_stp.gff
done
```

## Section 3.3-Obtain candidate RXLRs from all coding genes
### Step 1 
### Bash code chunk
```{bash eval = TRUE}
# In Bash environment
# Same code as Section 3.1 Step 7 to name variables and start loop
eval "$(conda shell.bash hook)"
conda activate agat

genome="data/PR-102_v4.fasta"
isolate=$(basename -s ".fasta" ${genome})
outdir="output_data"
assembly_dir="data"
rxlr_prefix="orfs_cand_RXLRs_"

# Run on all genomes being annotated
for orf_rxlr_list in $(ls -1 $outdir/${rxlr_prefix}*.txt); do
  # Establish output file name conventions
  list_name=$(basename --suffix ".txt" $orf_rxlr_list)
  iso_name=$(echo $list_name | sed "s/$rxlr_prefix//")
  echo $iso_name
  assembly=$(ls ${assembly_dir}/${iso_name}.fasta) || \
  assembly=$(ls ${assembly_dir}/other_refs/${iso_name}.fasta)
  
  ref_gff=$(ls $assembly_dir/${iso_name}.gff) || ref_gff=$(ls $assembly_dir/other_refs/${iso_name}.gff)
  echo $ref_gff

  # Re-name gene features
  agat_sp_manage_IDs.pl --tair --type_dependent \
    --prefix "${iso_name}_rxlrORF" \
    --nb 1 --gff $outdir/${list_name}_longnames.str_stp.gff \
    --output $outdir/${list_name}_longnames.str_stp.re-IDs.gff
    
  ls $ref_gff || \
    printf "GFF not found, configure \$iso_name:$iso_name and \$ref_gff:$ref_gff\" && exit 2"
    echo $ref_gff
  
  python3 scripts/complement_gff.py -r $ref_gff \
    -s $outdir/${list_name}_longnames.str_stp.re-IDs.gff \
    -o $outdir/${list_name}_longnames.str.stp.re-IDs.add_rxlrs.gff \
    -l $outdir/${list_name}_longnames.str_stp.re-IDs.overlapping_ref.gff \
    --unsorted

  # Extract all proteins including RXLR ORFs supplemented
  agat_sp_extract_sequences.pl \
    --gff $outdir/${list_name}_longnames.str.stp.re-IDs.add_rxlrs.gff \
    --fasta $assembly \
    --protein \
    -o $outdir/${list_name}_longnames.str_stp.re-IDs.add_rxlrs.fasta 
done
```

### Step 2, Part 1 of Step 3
### Bash code chunk
```{bash eval = TRUE}
# In bash environment
# Change directory names if needed
#In R bash code chunk, had to redefine 'genome' and 'isolate' variables
genome="data/Paga_3770v2_chr10.fasta"
isolate=$(basename -s ".fasta" ${genome})

peps_name="orfs_cand_RXLRs_${isolate}_longnames.str_stp.re-IDs.add_rxlrs" 
peps="output_data/${peps_name}.fasta" 
threads=10
python2.7 scripts/rxlr_signalpeptide.py $peps $threads secretome output_data/${peps_name}.rxlr_signalpep.out

tmhmm -short $peps > output_data/${peps_name}.tmhmm.out

# In bash environment, change directory names if needed 
hmmsearch --cpu 4 --seed 123 -T 0 --tblout output_data/${peps_name}.rxlr_WYfold_hmm.out data/WY_fold.hmm $peps > output_data/${peps_name}.rxlr_WYfold_hmm.log 
#remove slash before $1, when working with R bash code chunk  
egrep -v '^#' output_data/${peps_name}.rxlr_WYfold_hmm.out | awk '{print $1}' > output_data/${peps_name}.rxlr_WYfold_hmm.list 
```

### Step 3, part 2
### R code chunk
```{r eval = TRUE}
# In R environment 
wy_list_re_allgenes <- "orfs_cand_RXLRs_.*add_rxlrs.rxlr_WYfold_hmm.list" 
allgenes_wys <- list.files(path = "output_data/", 
                           pattern = wy_list_re_allgenes, 
                                     full.names = TRUE) %>% 
  map_dfr(read_wy_list_allgenes) %>% 
  unite("ID_isolate", ID, isolate, sep=":", remove = FALSE) %>% 
  mutate("method" = "WY_hmm") 
```

### Step 4
### R code chunk
```{r eval = TRUE}
# In R environment 
# SignalP results are to check if TM is before | after the SP 
sp3_fs_re_allgenes <- "*_longnames.str_stp.re-IDs.add_rxlrs.rxlr_signalpep.outsp3_tabular.tmp"
sp3_fs_allgenes <- list.files(path = "output_data", 
                                pattern = sp3_fs_re_allgenes, full.names = TRUE) 
tmm_fs_re_allgenes <- ".*orfs_cand_RXLRs_.*add_rxlrs.tmhmm.out" 
tmm_fs_allgenes <- list.files(path= "output_data", 
                     pattern = tmm_fs_re_allgenes, full.names = TRUE) 
 
# Dataframe for mapping, make sure filenames line up between SP and TM 
sp3_tmm_fs_allgenes <- tibble("sp3_fs" = sp3_fs_allgenes, 
                     "tmm_fs" = tmm_fs_allgenes) %>% 
  mutate("isolate_s" = str_extract(sp3_fs, 
                         "(?<=/orfs_cand_RXLRs_).*(?=_longnames)")) %>% 
  mutate("isolate_t" = str_extract(tmm_fs, 
                         "(?<=/orfs_cand_RXLRs_).*(?=_longnames)")) 
 
# Perform filtering function 
sp3_tmm_pass_allgenes <- pmap(sp3_tmm_fs_allgenes,~ read_prune_tms( 
                                                      ..1, ..2, ..3), 
                                              .id = 'isolate') %>% 
  bind_rows() %>% 
  unite("ID_isolate", Protein, isolate, sep=":", remove = FALSE) 
```

### Step 5
### R code chunk
```{r eval = TRUE}
# In R environment 
# Sequence filenames 
seqs_fs_re_allgenes <- "orfs_cand_RXLRs_.*_longnames.str_stp.re-IDs.add_rxlrs.fasta" 
seqs_fs_allgenes <- list.files(path = "output_data", 
                               pattern = seqs_fs_re_allgenes, 
                               full.names = TRUE) %>% 
  tibble("seqs_fs" = .) %>% 
  separate(seqs_fs, into = c("isolate", NA), 
           sep = "(?=_longnames)", remove = FALSE) %>% 
  mutate(isolate = str_remove(isolate, ".*orfs_cand_RXLRs_")) 
 
# SignalP filenames and join to seq fnames 
seqs_sp3_fs_allgenes <- sp3_fs_allgenes %>% 
  tibble("sp3_fs" = .) %>% 
  separate(sp3_fs, into = c("isolate", NA), 
           sep = "(?=_longnames)", remove = FALSE) %>% 
  mutate(isolate = str_remove(isolate, ".*orfs_cand_RXLRs_")) %>% 
  full_join(seqs_fs_allgenes, by = "isolate") %>% 
  select(seqs_fs, sp3_fs, isolate) 
 
# Run motif searches 
allgenes_re_effectr <- map2(seqs_sp3_fs_allgenes$seqs_fs, seqs_sp3_fs_allgenes$sp3_fs, effectr_eer_sp3) 
allgenes_re_effectr_tb <- tibble(bind_rows(allgenes_re_effectr)) %>% 
  unite("ID_isolate", ID, isolate, sep=":", remove = FALSE)
```

### Step 6
### R code chunk
```{r eval = TRUE}
# In R environment 

allgenes_re_effectr_tb_meths <- allgenes_re_effectr_tb %>%   
  bind_rows(allgenes_wys) %>%   
  group_by(ID_isolate, ID, isolate) %>%   
  summarize(methods_list = paste(method, collapse = ",")) %>%   
  filter((grepl("Whis2007", methods_list)) |   
         (grepl("Whis_rxlr", methods_list) & (grepl("Whis_eer", methods_list) | grepl("WY_hmm", methods_list))) |   
         (grepl("Win2007", methods_list) & (grepl("Whis_eer", methods_list) |  grepl("WY_hmm", methods_list)))) %>% 
  mutate(isolate = str_remove(isolate, "/")) 
```

### Step 7
### R code chunk
```{r eval = TRUE}
# In R environment 
outdir <- "output_data" 
out_prefix <- "/allgenes_RXLRs_" 
# Write each group as its own txt file 
allgenes_re_effectr_tb_meths %>% 
  filter(ID_isolate %in% sp3_tmm_pass_allgenes$ID_isolate) %>% 
  group_by(isolate) %>% 
  distinct(ID_isolate, .keep_all = TRUE) %>% 
 group_walk(~ writeLines(.x$ID, 
                         paste0(outdir, out_prefix, .y$isolate, ".txt"))) 
```
